---
title: "ReproducibleResearch_PA1"
author: "AMT"
date: "4/28/2022"
output: html_document
---

Reading and exploring the basics of the data

```{r, echo = TRUE}
unzip("./activity.zip")
activitydata <- read.csv("./activity.csv")
summary(activitydata)
dim(activitydata)
names(activitydata)
head(activitydata)
```

Transforming the date using lubridate

```{r, echo = TRUE}
library(lubridate)
activitydata$date <- ymd(activitydata$date)
```

Histogram of the total number of steps taken each day.

```{r, echo = TRUE}
library(ggplot2)
stepsperday <- aggregate(steps ~ date, activitydata, sum, na.rm = TRUE)

ggplot(stepsperday, aes(x=date, y=steps)) + geom_bar(stat = "identity") + xlab("Date") + ylab("Total Steps") + ggtitle("Total Steps per Date") + theme(plot.title = element_text(hjust = 0.5))

```

Mean and median of the total number of steps taken per date

Mean: 
```{r, echo = TRUE}
meansteps <- mean(stepsperday$steps)
meansteps
```

Median:
```{r, echo = TRUE}
mediansteps <- median(stepsperday$steps)
mediansteps
```

Histogram of the mean number of steps taken per day

```{r, echo = TRUE}
stepsmean <- aggregate(steps ~ date, activitydata, mean, na.rm = TRUE)

ggplot(stepsmean, aes(x=date, y=steps)) + geom_bar(stat = "identity") + xlab("Date") + ylab("Mean of Steps Taken per Day") + ggtitle("Mean Steps per Date") + theme(plot.title = element_text(hjust = 0.5))

```

Time series plot of the average number of steps taken (average daily activity pattern)

```{r, echo = TRUE}
activitydf <- data.frame(activitydata)
stepsperinterval <- aggregate(steps ~ interval, activitydf, mean, na.rm = TRUE)

plot(steps ~ interval, data = stepsperinterval, type = "l")

```

5-minute interval with the maximum number of steps

```{r, echo = TRUE}
intervalwithmaxsteps <- stepsperinterval[which.max(stepsperinterval$steps),]$interval
intervalwithmaxsteps
```

Imputing missing values

Total number of missing values:
```{r, echo = TRUE}
totalmissing <- sum(is.na(activitydata$steps))
totalmissing
```

Strategy for filling out missing value: substitute the missing values with means

```{r, echo = TRUE}
getmeansperinterval <- function(interval) {stepsperinterval[stepsperinterval$interval == interval,]$steps}
```

New dataset with imputed missing values

```{r, echo = TRUE}
activitydfimputed <- activitydf; 
  for(i in 1:nrow(activitydfimputed)) { 
    if(is.na(activitydfimputed[i,]$steps)){activitydfimputed[i,]$steps <- getmeansperinterval(activitydfimputed[i,]$interval)}}
```

Histogram of the total number of steps taken each day after missing values are imputed

```{r, echo = TRUE}
stepsperdayimputed <- data.frame(aggregate(steps ~ date, data = activitydfimputed, sum))

ggplot(stepsperdayimputed, aes(x=date, y=steps)) + geom_bar(stat = "identity") + xlab("Date") + ylab("Steps Taken per Day") + ggtitle("Total Steps per Date") + theme(plot.title = element_text(hjust = 0.5))

```

Mean and median of the total number of steps taken per date (based on the new dataset)

Mean:
```{r, echo = TRUE}
meanstepsimputed <- mean(stepsperdayimputed$steps)
meanstepsimputed
```

Median:
```{r, echo = TRUE}
medianstepsimputed <- median(stepsperday$steps)
medianstepsimputed
```

Activity patterns between weekdays and weekends

Creating a new factor variable in the dataset with two levels - "weekday" and "weekend"

```{r, echo = TRUE}
activitydfimputed$date <- as.Date(strptime(activitydfimputed$date, format = "%Y-%m-%d"))

activitydfimputed$day <- weekdays(activitydfimputed$date);
  for (i in 1:nrow(activitydfimputed)) {
    if (activitydfimputed[i,]$day %in% c("Saturday", "Sunday")) {
      activitydfimputed[i,]$day <- "weekend"
    }
    else{
      activitydfimputed[i,]$day <- "weekday"
    }
  }

stepsbyday <- aggregate(activitydfimputed$steps ~ activitydfimputed$interval + activitydfimputed$day, activitydfimputed, mean)
```

Panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.

```{r, echo = TRUE}
names(stepsbyday) <- c("interval", "day", "steps")
library(lattice)

xyplot(steps ~ interval | day, stepsbyday, type = "l", layout = c(1,2), xlab = "Interval", ylab = "Number of steps")

```

